Designing Types for R, Empirically

### 2020-10-31 paper/main.tex
edits

\documentclass[acmsmall, screen]{acmart}
\setlength{\pdfpagewidth}{8.5in}
\setlength{\pdfpageheight}{11in}
%\settopmatter{printfolios=true,printccs=true,printacmref=true}
\setcopyright{rightsretained}
\acmPrice{}
\acmDOI{10.1145/3360579}
\acmYear{2020}
\copyrightyear{2020}
\acmJournal{PACMPL}
\acmVolume{3}
\acmNumber{OOPSLA}
\acmArticle{153}
\acmMonth{10}

@@

\newcommand{\nul}{\texttt{\textbf{null}}\xspace}
\newcommand{\scalar}{\texttt{\textbf{scalar}}\xspace}
\renewcommand{\vector}{\texttt{\textbf{vector}}\xspace}
\newcommand{\class}{\texttt{\textbf{class}}\xspace}
\newcommand{\any}{\texttt{\textbf{any}}\xspace}
\newcommand{\environment}{\texttt{\textbf{env}}\xspace}
\newcommand{\expression}{\texttt{\textbf{expression}}\xspace}
\newcommand{\Language}{\texttt{\textbf{lang}}\xspace} %% Upper case!
\newcommand{\externalptr}{\texttt{\textbf{externalptr}}\xspace}
\renewcommand{\symbol}{\texttt{\textbf{symbol}}\xspace}
\newcommand{\pairlist}{\texttt{\textbf{pairlist}}\xspace}
\newcommand{\weakref}{\texttt{\textbf{weakref}}\xspace}
\renewcommand{\int}{\texttt{\textbf{int}}\xspace}
\newcommand{\chr}{\texttt{\textbf{chr}}\xspace}
\newcommand{\dbl}{\texttt{\textbf{dbl}}\xspace}
\newcommand{\lgl}{\texttt{\textbf{lgl}}\xspace}
\newcommand{\clx}{\texttt{\textbf{clx}}\xspace}
\newcommand{\raw}{\texttt{\textbf{raw}}\xspace}
\newcommand{\NA}{\texttt{{NA}}\xspace}
\newcommand{\VARGS}{\texttt{\textbf{...}}\xspace}
\newcommand{\T}{\ensuremath{T}\xspace}
\renewcommand{\S}{\ensuremath{S}\xspace}
\newcommand{\V}{\ensuremath{V}\xspace}
\newcommand{\A}{\ensuremath{A}\xspace}
\newcommand{\ID}{\ensuremath{ID}\xspace}
\newcommand{\F}{\ensuremath{F}\xspace}
\newcommand{\B}{\ensuremath{B}\xspace}
\newcommand{\FUN}[2]{\ensuremath{\langle #1 \rangle \rightarrow #2}\xspace}
\newcommand{\STRUCT}[1]{\texttt{\textbf{struct}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\LIST}[1]{\texttt{\textbf{list}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\CLASS}[1]{\texttt{\textbf{class}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\VEC}[1]{#1\k{[]}}
\newcommand{\NAVEC}[1]{\k{\string^}\!#1\k{[]}}
\newcommand{\NASCA}[1]{\k{\string^}\!#1\xspace}

@@

In order to keep the type signatures compact, we will compact union of
arrows into a single top-level arrow with unions at each argument
position. Thus, the shape of function signatures will be as follows:

@@

\noindent
In other words, we take the union of the types occurring at individual
argument positions rather than an union of function
types. Furthermore, we apply some transformation on the types to keep
the size of types in check.  Figure~\ref{fig:siplification-rules}
overviews the main simplification rules that we have adopted here.

@@

Assuming that type sequences can be reordered freely, we rewrite types
to minimize their size by removing redundant types, types that are
subsumed by subtyping, immutable lists, and remove \nul type to
replace them with nullables.  Higher-order functions are
conservatively treated as \any $\rightarrow$ \any.

It is noteworthy that by performing this compaction we are loosing
precision, that is to say, the synthesized signatures will suggest
that the function accepts combination of types that were not observed.

@@

\paragraph{Types}
The type information output by the tool includes the \emph{type tag}
of each value. Internal types are translated to names in the proposed
type system. The next bit of information is the \emph{class}, an
optional list of names that may be absent, and, in some cases, is
implicit (i.e. the interpreter blesses some values with the \k{matrix}
and \k{array} classes even without attributes).  Depending on a
value's type, the tool collects further information: (a) for vectors,
the presence of \NA values, (b) for lists, element types by a
recursive traversal, and (c) for promises, an approximation of the
expression type.  To obtain these types, we make use of R's C FFI and
use low-level machinery to collect information from the R run-time.
Types are completed during post-processing, and rely on the detailed
information made available by these reflection mechanisms.

@@

\paragraph{Missing arguments}
Parameters which receive no values when the function is called are
termed missing (not to be confused with \NA). This occurs when a
function is called with too few arguments and no default values are
specified for those missing arguments.  We record a \k{missing} type
for such argument.  There are two obvious ways to deal with missing
arguments: type them as \any or type them as some unit type.  We
conservatively type them as \any.

@@

@@

We have downloaded and installed all available CRAN packages. Out of
the \AllCranRnd packages, we managed to install \AllLoadableRnd. The
main reason for this is that \rdt (and, by extension, \typetracer) is
based on GNU-R 3.5.0 and some of the packages are not compatible with
this version.  Some packages also require extra native dependencies
which were not present on our servers. We defined two criteria for
including a package into the corpus:
\begin{inparaenum}[(1)]
  \item the package must have a runnable code that covers a
    significant part of the package source code from which type
    signatures could be inferred, and
  \item the package must have some reverse dependencies that will
    allows us to evaluate the inferred types using the runnable code
    from these dependencies.
\end{inparaenum}
The concrete thresholds used were: at least \ThresholdCodeCoverage of
expression coverage and at minimum \ThresholdRevdeps reverse
dependencies. The code coverage was computed for each package using
\covr\footnote{\url{https://github.com/r-lib/covr}}, the R code
coverage R tool.  The reverse package dependencies were extracted from
the package metadata using built-in functions.

@@

\paragraph{User code}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
To represent end-user code in the corpus, we turned to Kaggle, an
online platform for data-science and machine-learning. The website
allows people to share data-science competitions and data-analysis
problems together with data for which users try to find the best
solution (something like a repository of hackathon or datathon
code). The solutions, called \emph{kernels}, are then posted on Kaggle
either as plain scripts or as notebooks. One of the most popular
competitions is about predicting passenger survival on
Titanic\footnote{\url{https://www.kaggle.com/c/titanic}} with
\KaggleDownloaded kernels in R (over 1/4 of all available R kernels)
which we used for our corpus.

Unlike CRAN, Kaggle is not a curated repository and therefore there
are no guarantees about the quality of the code. After downloading all
of the \KaggleDownloaded kernels and extracting the R code from the
various formats,\footnote{We use {\sf rmarkdown} to convert from
notebooks to R.} we found that \KaggleDuplicated were whole-file
duplicates (\KaggleDuplicatedRatio). From the resulting
\KaggleRunnable kernels, \KaggleFail failed to execute. Next to
various runtime exceptions, common problems were missing libraries (no
longer available for R 3.5), parse errors and misspelled package
names. The final set contains \KaggleSuccess kernels with
\KaggleSuccessCodeRnd lines of R code. The Kaggle kernels are used for
additional validation of the inferred types.

\paragraph{Type usage}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
During execution 3,147 different types were observed.  Classes are the
most common types, accounting for roughly 31\% of types of arguments.
The most common classes are \c{matrices} (12\%), \c{data.frames}
(7.5\%), \c{formulas} (2\%), \c{factors} (2\%), and \c{tibbles} (2\%).
Roughly 25\% of classes are part of R's base libraries, the others are
user-defined.  Scalars and vectors are the next most common kind,
making up 41\% of remaining types. with scalars making up 28\% of
types and vectors 12\%.  Nulls and lists follow at 8\% and 7\%
respectively, and the vararg type makes up 6\% of arguments.  This all
totals up to over 90\% of types.  Table~\ref{ctb} reports on the 10
most frequent types occurring in the corpus. The first row of the
table reads: the \dbl type occurs in 12,298 (11.24\%) argument types,
and accounts for over 12 million (20.3\%) of the types observed by
\typetracer's dynamic analysis.

@@

This section attempts to evaluate how well the proposed type system is able
to describe the actual type signatures of functions. For this we focus on
how often there is a single type for a particular argument; this is because
union types and \any are less accurate (and would likely require a more
refined notion of subtyping or parametric polymorphism). Then, we evaluate
how robust the inferred signatures are by checking that they remain valid
for other inputs.  Lastly, we try to see if the current proposal would be
useful to programmers by allowing them to remove ad hoc checks and providing
useful documentation.

@@

The first part of our evaluation attempts to shed light on how good a fit
our proposed type system is with respect to common programming patterns
occurring in widely used R libraries.

@@

First we look at the share of monomorphic arguments and function signatures.
Monomorphic in this context means that the type is not relying on \any or
including a union.  The import of monomorphism in this context is that it
means our type language can accurately capture an argument's type or a
function signature. We get to that number in two steps.
Fig.~\ref{fig:unionfreq} shows the number of inferred argument types and
their size (in terms of members of the union). The figure shows the most
functions do not require a union at all (\PercUnitypedPositions of arguments
do not have a union), and only \PercManytypedPositions positions have unions
with more than three members.

@@

Table~\ref{fig:singlefreq} provides a breakdown of types occurring in
arguments without a union. Scalar, class and vector are the most
common type categories. The shaded rows correspond to polymorphic
types.  When an argument's type is \nul, we say that the argument is
polymorphic due to a limitation in our analysis: In R, it is common
for programmers to include default values for arguments, and in many
cases this value happens to be {\tt NULL}.  Our type analysis will
report a {\tt null} type for these arguments if they are {\it never
  passed a value} during testing.  We interpret these instances of
\nul as polymorphic to capture that we cannot be sure of the actual
type.

@@

\paragraph{Structs}
While experimenting with various design, we consider adding a struct
type to capture lists with named elements that can be accessed with
the \code{$} operators. We ended up discarding those types as they
grew large and were often only representative of the example data
being manipulated.  Consider function \code{cv.model}, its argument
\code{x} is observed to be of \CLASS{\texttt{aov}, \texttt{lm}} or
\CLASS{\texttt{lm}}.  Internally, linear models are represented as
lists with named elements.  The pollution is illustrated by the lines
after the function definition.  They load an example data set and test
the function \code{cv.model}.  The \code{data(sweetpotato)} expression
loads a sample data set. The fields of \code{sweetpotato} will be
recorded when \code{cv.model} is called.

@@

\paragraph{Objects}
While we record classes, our analysis does not deal with method
dispatch.  R has multiple disparate object systems called S3, S4, and
R5.  The \code{class} attribute is used by these systems to dispatch
methods. S3 does single dispatch, S4 does multiple dispatch and R5
supports imperative objects.  The mechanics of S4 dispatch are more
complex than for S3, and users can define their own class hierarchies
that we would need to incorporate in our type analysis and contract
checking frameworks.  We found limited use of S4 during our
analysis. Coming up with a type system that accounts for all of these
factors and consolidates multiple object-orientation frameworks in a
single language design is an interesting problem in and of itself, one
we leave for future work.

@@

Finally, we point out that assertion failures of, e.g., \CLASS{{\tt
    data.frame}} values being passed to \CLASS{{\tt data.frame, tbl,
    tbl\_df}} arguments and \CLASS{{\tt xml\_node}} values being
passed to an argument expecting other XML-like classes are related to
our simplified take on class types.  Our type system does not encode
user-defined subtyping and coercion, which could help address these
mismatches.

@@

We see that even though the double vector and matrix issue was wildly
prevalent in the raw, dynamic contract evaluation numbers, the number
of actual function argument types that were violated is very small.
The story is similar with the double and integer coercion we mentioned
earlier: it represents many dynamic contract failures, but very few of
the \VEC\int-typed arguments have their contracts violated by
\VEC\dbl-typed values.  Row six is interesting: we see that rather
often arguments expecting \CLASS{\texttt{timeSeries}} data are passed
\CLASS{\texttt{matrix}} values.  This is a quirk of the
\code{timeSeries} package, whose functions often accept matrices and
vectors, converting them to time series in an ad hoc manner.  Note
that the code coverage of the \code{timeSeries} tests, examples, and
vignettes package code is only 58\%, which is one possible explanation
of why these contract failures are occurring: the types that
\typetracer generates are only as good as the test code its run on.

@@

Had we failed to capture some key usage pattern of R with our type
annotation framework, we would likely see it here, and we can see this
in action if we consider Table~\ref{tbl:fail-breakdown-3}, which was
obtained identically to Table~\ref{tbl:fail-breakdown-2} except
selecting arguments in the 80th percentile instead.  The most frequent
argument type violation pattern in that of \CLASS{{\tt matrix}},
\VEC\dbl, and \CLASS{\texttt{data.frame}} values passed to arguments
expecting \CLASS{{\tt timeSeries}}.  This occurs in 35.90\% of such
arguments, and represents cases where tests did not adequately cover
all valid function inputs.  Separate from the issue of testing, we can
capture this behaviour with user-defined subtyping or coercion, as the
data types which were passed are readily convertible to \CLASS{{\tt
    timeSeries}}.

@@

  BBmisc::strrepeat("a", 3) # => "aaa"
  BBmisc::strrepeat(c("a", "b"), 3) # => "ababab"
 \end{lstlisting}
 This function was only ever tested with unit-length vectors passed to
 {\tt x}, even though technically it can handle longer vectors, as per
 the two sample calls above.  This could be attributed to poor quality
 testing, or misunderstanding language semantics (e.g.,
 misunderstanding the semantics of \code{paste0} and \code{rep.int}),
 but we found other instances of type errors where the functions
 really ought to have been tested with the offending type, for
 instance:
 \begin{lstlisting}
  combinat::permn <- function (x, fun = NULL, ...) {
    if (is.numeric(x) && length(x) == 1 && x > 0 && trunc(x) == x) 
        x <- seq(x)
    # ...
  }
 \end{lstlisting}
 As per the documentation, this function is intended to take a vector,
 and generate all permutations of elements in that vector.  If given a
 scalar ``integer'' $n$, it will generate all permutations for the
 list $[1, 2, ..., n]$.  Interestingly, the function was only tested
 by passing an integer, and not ever with a vector (even though,
 presumably, that is the main utility of the function).

@@

 By-and-large, we saw no meaningful difference in the two data sets,
 with the contract assertion failure patterns being repeated from the
 reverse dependencies.  Overall, we observed that
 \PERCFAILEDASSERTIONSKAGGLE of all contract assertions failed while
 running Kaggle code.  If we remove assertion failures related to our
 simplifying assumption that function types will not have more than 20
 arguments, that number drops to a mere
 \PERCFAILEDASSERTIONSKAGGLENOUNDEF.  In all,
 \PROPFUNSWITHFAILEDASSERTKAGGLE functions had at least one contract
 failure.  There were \NUMTOTALASSERTIONSKAGGLE assertions in total,
 on \NUMASSERTEDFUNCTIONSKAGGLE functions.

@@

 Another interesting entry in Table~\ref{tbl:fail-breakdown-0-kaggle}
 is the fifth row, where a \CLASS{{\tt ixforeach, iter}} is being
 passed to an argument expecting a long union of classes, each of the
 form \CLASS{{\tt X, iter}}.  This is likely an instance where the
 user defined their own class, {\tt ixforeach}, and wanted to use the
 \code{iterators} package (the user called \code{iterator::iter} with
 a \CLASS{{\tt ixforeach}}, and it gained the class {\tt iter} on
 return).  As mentioned, accounting for object-orientation in the
 type system is beyond the scope of this work, and such an inclusion
 would allow us to better type situations like this.

@@

There are a number of ways to check the types of function parameters
in R.  The default and most common way\footnote{\CranStopifnotRatio of
all runtime checks in the whole of CRAN} is to use the
\code{stopifnot} function from the R base package. It takes a number
of R expressions which all should evaluate to true otherwise a runtime
exception is thrown with a message quoting the corresponding failed
expression. For example, the following code checks whether a given
parameter \code{x} is a scalar string:
%
\begin{lstlisting}
stopifnot(is.character(x), length(x) == 1L, !is.na(x))
\end{lstlisting}

Besides \code{stopifnot}, there are 4 packages in
CRAN\footnote{Packages are available on the CRAN website:
\url{https://cran.r-project.org/web/packages/}} that focus on runtime
assertions: \emph{assertive}, \emph{ensurer}, \emph{assertr} and
\emph{assertthat}. \emph{assertive} and \emph{ensurer} have not been
updated since 2016 and 2015 respectively, and \emph{assertr} is used
by only \AssertrRevdeps other packages and currently focuses on
checking properties of data frames. Only the \emph{assertthat} package
is maintained and used (with \AssertthatRevdeps reverse
dependencies). The advantage of \emph{assertthat} over the R's default
is that it provides much better error messages.

One way to asses the usefulness of our type checking system is to find
out how many of the existing type checking constrains could be
replaced by \contractr.  To measure this, we have extracted all calls
to \code{stopifnot} and \code{assertthat} assertions and checked which
among them could be either completely replaced by \contractr or at
least partially simplified by removing a portion of an assertion
expression. This is useful, because a common pattern is that the first
part of the parameter assertion checks its type while the rest checks
its value. In the example above, the whole expression could be
replaced by a \chr type check.

>>>>

\documentclass[acmsmall, screen]{acmart}
\setlength{\pdfpagewidth}{8.5in}
\setlength{\pdfpageheight}{11in}
%\settopmatter{printfolios=true,printccs=true,printacmref=true}
\setcopyright{rightsretained} 
\acmJournal{PACMPL}
\acmYear{2020} \acmVolume{4} \acmNumber{OOPSLA} \acmArticle{181} \acmMonth{11} \acmPrice{}\acmDOI{10.1145/3428249}

@@

\newcommand{\nul}{\texttt{\textbf{null}}\xspace}
\newcommand{\scalar}{\texttt{\textbf{scalar}}\xspace}
\renewcommand{\vector}{\texttt{\textbf{vector}}\xspace}
\newcommand{\class}{\texttt{\textbf{class}}\xspace}
\newcommand{\any}{\texttt{\textbf{any}}\xspace}
\newcommand{\environment}{\texttt{\textbf{env}}\xspace}
\newcommand{\expression}{\texttt{\textbf{expression}}\xspace}
\newcommand{\Language}{\texttt{\textbf{lang}}\xspace} %% Upper case!
\newcommand{\externalptr}{\texttt{\textbf{externalptr}}\xspace}
\renewcommand{\symbol}{\texttt{\textbf{symbol}}\xspace}
\newcommand{\pairlist}{\texttt{\textbf{pairlist}}\xspace}
\newcommand{\weakref}{\texttt{\textbf{weakref}}\xspace}
\renewcommand{\int}{\texttt{\textbf{int}}\xspace}
\newcommand{\chr}{\texttt{\textbf{chr}}\xspace}
\newcommand{\dbl}{\texttt{\textbf{dbl}}\xspace}
\newcommand{\lgl}{\texttt{\textbf{lgl}}\xspace}
\newcommand{\clx}{\texttt{\textbf{clx}}\xspace}
\newcommand{\raw}{\texttt{\textbf{raw}}\xspace}
\newcommand{\NA}{\texttt{{NA}}\xspace}
\newcommand{\VARGS}{\texttt{\textbf{...}}\xspace}
\newcommand{\T}{\ensuremath{T}\xspace}
\renewcommand{\S}{\ensuremath{S}\xspace}
\newcommand{\V}{\ensuremath{V}\xspace}
\newcommand{\A}{\ensuremath{A}\xspace}
\newcommand{\ID}{\ensuremath{ID}\xspace}
\newcommand{\F}{\ensuremath{F}\xspace}
\newcommand{\B}{\ensuremath{B}\xspace}
\newcommand{\FUN}[2]{\ensuremath{\langle #1 \rangle \rightarrow #2}\xspace}
\newcommand{\STRUCT}[1]{\texttt{\textbf{struct}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\LIST}[1]{\texttt{\textbf{list}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\CLASS}[1]{\texttt{\textbf{class}}\ensuremath{\langle #1\rangle}}
\newcommand{\VEC}[1]{#1\k{[]}}
\newcommand{\NAVEC}[1]{\k{\string^}\!#1\k{[]}}
\newcommand{\NASCA}[1]{\k{\string^}\!#1\xspace}

@@

Later in this paper, we will introduce a tool to synthesize function type signatures from running R code, wherein each invocation of a function will generate a trace representing the types of function arguments and returns.
These traces will be combined into an overall arrow type for the function, and in order to keep these signatures compact we compact the traces into a single top-level arrow with unions at each argument
position. Thus, the shape of function signatures will be:

@@

\noindent
In other words, we take the union of the types occurring at individual
argument positions rather than an union of function
types. Furthermore, we apply some transformation on the types to keep
the size of types in check.  Figure~\ref{fig:siplification-rules}
overviews the main simplification rules we adopted.

@@

Assuming that type sequences can be reordered freely, we rewrite types
to minimize their size by removing redundant types, types that are
subsumed by subtyping, immutable lists, and replace \nul types with nullables.  Higher-order functions are
conservatively treated as \any $\rightarrow$ \any.

It is noteworthy that by performing this compaction we are losing
precision, and the synthesized signatures will suggest combinations of argument types that were not observed.

@@

\paragraph{Types}
The type information output by the tool includes the \emph{type tag}
of each value. Internal types are translated to names in the proposed
type language. The next bit of information is the \emph{class}, an
optional list of names that may be absent, and, in some cases, is
implicit (i.e. the interpreter blesses some values with the \k{matrix}
and \k{array} classes even without attributes).  Depending on a
value's type, the tool collects further information: (a) for vectors,
the presence of \NA values, (b) for lists, element types by a
recursive traversal, and (c) for promises, an approximation of the
expression type.  To obtain these types, we make use of R's C FFI and
use low-level machinery to collect information from the R run-time.
Types are completed during post-processing, and rely on the detailed
information made available by these reflection mechanisms.

@@

\paragraph{Missing arguments}
Parameters which receive no values when the function is called are
termed missing (not to be confused with \NA). This occurs when a
function is called with too few arguments and no default values are
specified for those missing arguments.  We record a \k{missing} type
for such arguments.  There are two obvious ways to deal with missing
arguments: type them as \any or type them as some unit type.  We
conservatively type them as \any.

@@

Our aim is to gather data with which we can validate our type language design.
To that end, we propose an experiment wherein we use \typetracer to synthesize types for some core corpus of packages, and validate those types by using \contractr and installing contracts on core corpus functions and running the client code of this corpus.

@@

We have downloaded and installed all available CRAN packages. Out of
the \AllCranRnd packages, we managed to install \AllLoadableRnd. The
main reason for this is that \rdt (and, by extension, \typetracer) is
based on GNU-R 3.5.0 and some of the packages are not compatible with
this version.  Some packages also require extra native dependencies
which were not present on our servers.

We defined two criteria for
including a package into the corpus:
\begin{inparaenum}[(1)]
  \item the package must have runnable code that covers a
    significant part of the package source code from which type
    signatures could be inferred, and
  \item the package must have some reverse dependencies that will
    allow us to evaluate the inferred types, using the runnable code
    from these dependencies.
\end{inparaenum}
The concrete thresholds used were: at least \ThresholdCodeCoverage of
expression coverage and at minimum \ThresholdRevdeps reverse
dependencies. The code coverage was computed for each package using
\covr\footnote{\url{https://github.com/r-lib/covr}}, the R code
coverage R tool.  The reverse package dependencies were extracted from
the package metadata using built-in functions.

@@

\paragraph{User code}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
To represent end-user code in the corpus, we turned to Kaggle, an
online platform for data science and machine learning. The website
allows people to share data science competitions and data analysis
problems together with data for which users try to find the best
solution (something like a repository of hackathon or datathon
code). The solutions, called \emph{kernels}, are then posted on Kaggle
either as plain scripts or as notebooks. One of the most popular
competitions is about predicting passenger survival on
Titanic\footnote{\url{https://www.kaggle.com/c/titanic}} with
\KaggleDownloaded kernels in R (over 1/4 of all available R kernels)
which we used for our corpus.

Unlike CRAN, Kaggle is not a curated repository and therefore there
are no guarantees about the quality of the code. After downloading all
of the \KaggleDownloaded kernels and extracting the R code from the
various formats,\footnote{We use {\sf rmarkdown} to convert from
notebooks to R.} we found that \KaggleDuplicated were whole-file
duplicates (\KaggleDuplicatedRatio). From the resulting
\KaggleRunnable kernels, \KaggleFail failed to execute. Next to
various runtime exceptions, common problems were missing libraries (no
longer available for R 3.5), parse errors, and misspelled package
names. The final set contains \KaggleSuccess kernels with
\KaggleSuccessCodeRnd lines of R code. The Kaggle kernels are used for
additional validation of the inferred types.

\paragraph{Type usage}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
During execution, 3,147 different types were observed.  Classes are the
most common types, accounting for roughly 31\% of types of arguments.
The most common classes are \c{matrices} (12\%), \c{data.frames}
(7.5\%), \c{formulas} (2\%), \c{factors} (2\%), and \c{tibbles} (2\%).
Roughly 25\% of classes are part of R's base libraries, the others are
user-defined.  Scalars and vectors are the next most common kind,
making up 41\% of remaining types. with scalars making up 28\% of
types and vectors 12\%.  Nulls and lists follow at 8\% and 7\%
respectively, and the vararg type makes up 6\% of arguments.  This all
totals up to over 90\% of types.  Table~\ref{ctb} reports on the 10
most frequent types occurring in the corpus. The first row of the
table reads: the \dbl type occurs in 12,298 (11.24\%) argument types,
and accounts for over 12 million (20.3\%) of the types observed by
\typetracer's dynamic analysis.

@@

This section attempts to evaluate how well the proposed type language is able
to describe the actual type signatures of functions. For this we focus on
how often there is a single type for a particular argument; this is because
union types and \any are less accurate (and would likely require a more
refined notion of subtyping or parametric polymorphism). Then, we evaluate
how robust the inferred signatures are by checking that they remain valid
for other inputs.  Lastly, we try to see if the current proposal would be
useful to programmers by allowing them to remove ad hoc checks and providing
useful documentation.

@@

The first part of our evaluation attempts to shed light on how good a fit
our proposed type language is with respect to common programming patterns
occurring in widely used R libraries.

@@

First we look at the share of monomorphic arguments and function signatures.
Monomorphic in this context means that the type is not relying on \any or
including a union.  The import of monomorphism in this context is that it
means our type language can accurately capture an argument's type or a
function's signature. We get to that number in two steps.
Fig.~\ref{fig:unionfreq} shows the number of inferred argument types and
their size (in terms of members of the union). The figure shows that most
functions do not require a union at all (\PercUnitypedPositions of arguments
do not have a union), and only \PercManytypedPositions of positions have unions
with more than three members.

@@

Table~\ref{fig:singlefreq} provides a breakdown of types occurring in
arguments without a union. Scalar, class and vector are the most
common type categories. The shaded rows correspond to polymorphic
types.  When an argument's type is inferred to be \nul, we say that the argument is
polymorphic due to a limitation in our analysis: In R, it is common
for programmers to include default values for arguments, and in many
cases this value happens to be {\tt NULL}.  Our type analysis will
report a \nul type for these arguments if they are {\it never
  passed a value} during testing.  We interpret these instances of
\nul as polymorphic to capture that we cannot be sure of the actual
type.

@@

\paragraph{Structs}
While experimenting with various design, we consider adding a struct
type to capture lists with named elements that can be accessed with
the \code{$} operators. We ended up discarding those types as they
grew large and were often only representative of the example data
being manipulated.  Consider function \code{cv.model}, its argument
\code{x} is observed to be of \CLASS{\texttt{aov}, \texttt{lm}} or
\CLASS{\texttt{lm}}.  Internally, linear models are represented as
lists with named elements.  The pollution is illustrated by the lines
after the function definition, where the \code{data(sweetpotato)} expression
loads a sample data set to test the function.  The fields of \code{sweetpotato} will be
recorded when \code{cv.model} is called.

@@

\paragraph{Objects}
While we record classes, our analysis does not deal with method
dispatch.  R has multiple disparate object systems called S3, S4, and
R5.  The \code{class} attribute is used by these systems to dispatch
methods. S3 does single dispatch, S4 does multiple dispatch, and R5
supports imperative objects.  The mechanics of S4 dispatch are more
complex than for S3, and users can define their own class hierarchies
that we would need to incorporate in our type analysis and contract
checking frameworks.  We found limited use of S4 during our
analysis. Coming up with a type system that accounts for all of these
factors and consolidates multiple object-orientation frameworks in a
single language design is an interesting problem in and of itself, one
we leave for future work.

@@

Finally, we point out that assertion failures of, e.g., \CLASS{{\tt
    data.frame}} values being passed to \CLASS{{\tt data.frame, tbl,
    tbl\_df}} arguments and \CLASS{{\tt xml\_node}} values being
passed to an argument expecting other XML-like classes are related to
our simplified take on class types.  Our type language does not encode
user-defined subtyping and coercion, which could help address these
mismatches.

@@

We see that even though the double vector and matrix issue was wildly
prevalent in the dynamic contract evaluation numbers, the number
of actual function argument types that were violated is very small.
The story is similar with the double and integer coercion we mentioned
earlier: it represents many dynamic contract failures, but very few of
the \VEC\int-typed arguments have their contracts violated by
\VEC\dbl-typed values.  Row six is interesting: we see that rather
often arguments expecting \CLASS{\texttt{timeSeries}} data are passed
\CLASS{\texttt{matrix}} values.  These are all from the
\code{timeSeries} package, whose functions often accept matrices and
vectors, converting them to time series in an ad hoc manner.  Note
that the code coverage of the \code{timeSeries} tests, examples, and
vignettes package code is only 58\%, which is one possible explanation
for these contract failures: the types that
\typetracer generates are only as good as the test code its run on.

@@

Had we failed to capture some key usage pattern of R with our type
annotation framework, we would likely see it here. For example, consider Table~\ref{tbl:fail-breakdown-3}, which was
obtained identically to Table~\ref{tbl:fail-breakdown-2} except
selecting arguments in the 80th percentile instead.  The most frequent
argument type violation pattern in that of \CLASS{{\tt matrix}},
\VEC\dbl, and \CLASS{\texttt{data.frame}} values passed to arguments
expecting \CLASS{{\tt timeSeries}}.  This occurs in 35.90\% of such
arguments, and represents cases where tests did not adequately cover
all valid function inputs.  Separate from the issue of testing, we can
capture this behaviour with user-defined subtyping or coercion, as the
data types which were passed are readily convertible to \CLASS{{\tt
    timeSeries}}.

@@

  BBmisc::strrepeat("a", 3) # => "aaa"
  BBmisc::strrepeat(c("a", "b"), 3) # => "ababab"
 \end{lstlisting}
 This function was only ever tested with unit-length vectors passed to
 {\tt x}, even though technically it can handle longer vectors, as per
 the two sample calls above.  This could be attributed to poor quality
 testing, or misunderstanding language semantics (e.g.,
 misunderstanding the semantics of \code{paste0} and \code{rep.int}),
 but we found other instances of type errors where the functions
 really ought to have been tested with the offending type, for
 instance:
 \begin{lstlisting}
  combinat::permn <- function (x, fun = NULL, ...) {
    if (is.numeric(x) && length(x) == 1 && x > 0 && trunc(x) == x) 
        x <- seq(x)
    # ...
  }
 \end{lstlisting}
 As per the documentation, this function is intended to take a vector,
 and generate all permutations of elements in that vector.  If given a
 scalar integer $n$, it will generate all permutations for the
 list $[1, 2, ..., n]$.  Interestingly, the function was only tested
 by passing an integer, and not ever with a vector (even though,
 presumably, that is the main utility of the function).

@@

 By-and-large, this evaluation did not reveal any new insights,
 with the contract assertion failure patterns being repeated from the
 reverse dependencies.  Overall, we observed that
 \PERCFAILEDASSERTIONSKAGGLE of all contract assertions failed while
 running Kaggle code.  If we remove assertion failures related to our
 simplifying assumption that function types will not have more than 20
 arguments, that number drops to a mere
 \PERCFAILEDASSERTIONSKAGGLENOUNDEF.  In all,
 \PROPFUNSWITHFAILEDASSERTKAGGLE functions had at least one contract
 failure.  There were \NUMTOTALASSERTIONSKAGGLE assertions in total,
 on \NUMASSERTEDFUNCTIONSKAGGLE functions.

@@

 Another interesting entry in Table~\ref{tbl:fail-breakdown-0-kaggle}
 is the fifth row, where a \CLASS{{\tt ixforeach, iter}} is being
 passed to an argument expecting a long union of classes, each of the
 form \CLASS{{\tt \_, iter}}.  This is likely an instance where the
 user defined their own class, {\tt ixforeach}, and wanted to use the
 \code{iterators} package (the user called \code{iterator::iter} with
 a \CLASS{{\tt ixforeach}}, and it gained the class {\tt iter} on
 return).  As mentioned, accounting for object-orientation in the
 type system is beyond the scope of this work, and such an inclusion
 would allow us to better type situations like this.

@@

There are a number of ways to check the types of function parameters
in R.  The default and most common way\footnote{\CranStopifnotRatio of
all runtime checks in the whole of CRAN} is to use the
\code{stopifnot} function from the R base package. It takes a number
of R expressions which should all evaluate to \code{TRUE}, otherwise a runtime
exception is thrown with a message quoting the failed
expression. For example, the following code checks whether a given
parameter \code{x} is a scalar string:
%
\begin{lstlisting}
stopifnot(is.character(x), length(x) == 1L, !is.na(x))
\end{lstlisting}

Beyond \code{stopifnot}, there are 4 packages in
CRAN\footnote{Packages are available on the CRAN website:
\url{https://cran.r-project.org/web/packages/}} that focus on runtime
assertions: \emph{assertive}, \emph{ensurer}, \emph{assertr}, and
\emph{assertthat}. \emph{assertive} and \emph{ensurer} have not been
updated since 2016 and 2015 respectively, and \emph{assertr} is used
by only \AssertrRevdeps other packages and currently focuses on
checking properties of data frames. Only \emph{assertthat} is maintained and used (with \AssertthatRevdeps reverse
dependencies). The advantage of \emph{assertthat} over the R default
is that it provides much better error messages.

One way to assess the usefulness of our type checking framework is to see how many existing user-defined type checks can be replaced by \contractr.
To measure this, we extracted all calls
to \code{stopifnot} and \code{assertthat} assertions, and checked which
among them could either be completely replaced by \contractr, or at
least partially simplified by removing a portion of an assertion
expression. Partial simplification is useful, as a common pattern in this ad hoc type checking is that the first
part of the assertion checks the value's type, while the rest checks more detailed properties. In the example above, the whole expression could be
replaced by a \chr type check.

Out of the \CorpusLoadable packages, \CorpusAssertsInPackages use
runtime assertions. Altogether, there are \CorpusAsserts assertions in
\CorpusAssertsInFunctions functions. Among these, \contractr can
replace \CorpusTypedAsserts (\CorpusTypedAssertsRatio) assertion calls
across \CorpusTypedAssertsPackages packages and
\CorpusTypedAssertsFunctions functions.  Furthermore, an additional
\CorpusPartiallyTypedAsserts (\CorpusPartiallyTypedAssertsRatio)
asserts in \CorpusPartiallyTypedAssertsPackages packages and
\CorpusPartiallyTypedAssertsFunctions functions could have been
simplified.

Checking the type of function parameters is not something that is seen
often in the R code. In the whole of CRAN, there are only
\CranAssertsRnd asserts in \CranAssertsInFunctionsRnd functions defined
in \CranAssertsInPackagesRnd packages.  One may speculate that this is
the case due to the verbosity and inconvenience of the existing
assertion tools.  Our system can infer type annotations for existing
functions automatically. This can remove or partially remove over
\CorpusPartiallyTypedAssertsRatio of existing assertions.
